# Comparing two means

## Checking data for the independent samples t-test

### Producing histograms and boxplots by a second variable

We can create histograms and boxplots separated by a second variable in (at least) two ways: using Base R or `ggplot2` graphics. We will demonstrate using the birthweight data in `Example_5.1.dta`.

```{r}
library(haven)
library(labelled)
library(skimr)
library(ggplot2)
library(jmv)

bwt <- read_dta("data/examples/Example_5.1.dta") %>% 
  unlabelled()

skim(bwt)
summary(bwt$gender)
```

To use Base R graphics, we can create subsets of the birthweight data, subsetted for males and females separately. Note here that `gender` is a factor, so we need to select based on the factor labels, not the underlying numeric code.

```{r}
bwt_m <- subset(bwt, bwt$gender=="Male")
bwt_f <- subset(bwt, bwt$gender=="Female")
```

We can now create hisotgrams and boxplots for males and females separately, in the usual way. To place the graphs next to each other in a single figure, we can use the `par` function. The `par` function sets the graphics parameters. Essentially, we want to tell R to split a plot window into a matrix with *nr* rows and *nc* columns, and we can decide to fill the cells by rows (`mfrow`) or columns (`mfcols`). For example, to plot four figures in a single plot, filled by rows, we use `par(mfrow=c(2,2))`.

When we are done plotting multiple graphs, we can reset the plot window by submitting `par(mfrow=c(1,1))`.

```{r}
par(mfrow=c(2,2))
hist(bwt_m$birthweight, xlim=c(2.5, 4.5), xlab="Birthweight (kg)", main="Males")
hist(bwt_f$birthweight, xlim=c(2.5, 4.5), xlab="Birthweight (kg)", main="Females")

boxplot(bwt_m$birthweight, ylim=c(2.5, 4.5), ylab="Birthweight (kg)", main="Males")
boxplot(bwt_f$birthweight, ylim=c(2.5, 4.5), ylab="Birthweight (kg)", main="Females")
par(mfrow=c(1,1))
```

To produce separate histograms in `ggplot2`, we use the `facet_wrap` function to create a grid of plots. We can define the variable(s) to be plotted by in the `vars()`, and optionally, the number of rows (`nrow=`) and number of columns (`ncol=`).

```{r}
# Overall histogram of birthweight
ggplot(bwt, aes(x=birthweight)) + 
  geom_histogram(breaks=seq(2.5, 4.5, 0.25), colour="black", fill="grey") + 
  labs(x="Birthweight (kg)", y="Frequency") +
  theme_classic()

# Histogram by gender
ggplot(bwt, aes(x=birthweight)) + 
  geom_histogram(breaks=seq(2.5, 4.5, 0.25), colour="black", fill="grey") + 
  facet_wrap(vars(gender), nrow=1, ncol=2) +
  labs(x="Birthweight (kg)", y="Frequency") +
  theme_classic()
```

While it is possible to use `facet_wrap` to produce separate boxplots, we can use the fact that the boxplot allows an `x` variable to be assigned to the ggplot aesthetic. By defining `birthweight` as the `y` variable and `gender` as the `x` variable, we can produce two boxplots in the same figure:

```{r}
ggplot(bwt, aes(x=gender, y=birthweight)) +
  geom_boxplot() +
  labs(y="Birthweight (kg)", x="Gender") +
  theme_classic()
```


### Producing split summary statistics

The `descriptives` function within the `jmv` function allows summary statistics to be calculated within subgroups using the `splitBy` argument:

```{r}
descriptives(data=bwt, vars=birthweight, splitBy=gender)
```

## Independent samples t-test

```{r}
ttestIS(data=bwt, vars=birthweight, group=gender)
ttestIS(data=bwt, vars=birthweight, group=gender, meanDiff=TRUE, ci=TRUE)
ttestIS(data=bwt, vars=birthweight, group=gender, meanDiff=TRUE, ci=TRUE, welchs=TRUE)
```


## Checking the assumptions for a Paired t-test

Before performing a paired t-test, you must check that the assumptions for the test have been met. Using the dataset `Example_5.2.dta` to show that the difference between the pair of measurements between the sites is normally distributed, we first need to compute a new variable of the differences and examine its histogram.

```{r}
sbp <- read_dta("/Users/td/Documents/GithubRepos/phcm9795/data/examples/Example_5.2.dta")
sbp$diff = sbp$sbp_dp - sbp$sbp_tp
hist(sbp$diff, xlab="Blood pressure (mmHg)", main="Difference in systolic blood pressure")
```

 
## Paired t-Test

To perform a paired t-test we will use the dataset `Example_5.2.dta`.

```{r}
ttestPS(data=sbp, pairs=list(list(i1 = 'sbp_dp', i2 = 'sbp_tp')), meanDiff=TRUE, ci=TRUE, norm=TRUE)
```

